Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-08 paper/manuscript.tex
edits to all sections up to results

We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
We used the T332N variant of BG505 Env because it has a glycosylation site that is targeted by many anti-HIV antibodies but is absent in wildtype BG505~\citep{sanders2013next}.
The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across the clade A alignment (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning approach in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed summaries of all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to the earlier inferior LAI dataset. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
Direct comparisons between experimentally measured amino-acid preferences and amino-acid frequencies in natural sequences are confounded by the fact that the natural sequences are evolutionarily related.
This problem can be overcome by making the comparison in a phylogenetic context.
Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms,bloom2017identification}.
Importantly, since we expect many sites in Env to be under diversifying selection from immunity, we extended the ExpCM's to draw the relative dN/dS parameter ($\omega$) from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.

@@

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The upper overlaid color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower overlaid color bar indicates the evidence that the site is evolving faster or slower than expected in nature given the experiments~\citep[see][]{bloom2017identification}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in Table \ref{tab:phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as Figure~\ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

The overall distribution of shifts between BG505 and BF520 is shown in Figure~\ref{fig:prefsdist}B.
Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
How should we interpret this distribution -- have mutational effects have shifted a lot, or not very much?
We can establish an upper-bound for how much sites can shift by comparing Env to a \emph{non}-homologous protein.
Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
We can also establish a lower-bound by creating a null distribution for the expected shifts if all differences are simply due to experimental noise.
This null distribution is created by randomizing the experimental replicates among Envs.
Figure~\ref{fig:prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.

@@

However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
Several other sites show similar but somewhat milder flips in which amino acids they prefer.
In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to fully rationalize many aspects of the observed sequence-function relationships.

@@

What distinguishes the sites that have undergone significant shifts in their amino-acid preferences?
Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most of the shifted sites have not substituted (only 8 of the 30 shifted sites differ between the two Envs).
The lack of a strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which have found that shifts are dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

\subsection{Comparing selection in the lab to natural selection}
All the foregoing experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
For instance, antibody pressure might increase the rate of evolution at some sites~\citep{albert1990rapid,wei2003antibody,richman2003rapid}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites. 
Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.
%In contrast, standard methods for analyzing protein evolution that do not use experiments only test if sites are changing faster or slower than expected under neutral selection~\citep[so-called dN/dS tests;][]{yang2000codon,kosakovsky2005not,murrell2013fubar}.
% -- something that is of less utility since we already know that Env is under strong functional constraint.

@@

We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520. 
The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
As shown in Table~\ref{tab:phydms}, the standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution.
This finding is only of modest interest, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).

The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B.
The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.

What causes some sites to evolve faster or slower in nature than expected from the experiments?
The answer in both cases is likely to be immune selection.
Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
It is known that Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
Sites that contribute to such immune-evasion strategies will be under a constraint in nature that is not present in our experiments, since the experiments do not select for immune evasion.
%Indeed, many of the sites that evolve slower than expected appear to be involved in these general immune-evasion strategies.
For instance, sites of N-linked glycans tend to be more conserved in nature than the experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as are sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.

>>>>

We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 have typical divergence for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

We therefore modified the deep mutational scanning strategy of \citet{haddox2016experimental} to yield the approach in Figure~\ref{fig:dms_schematic}A.
This approach had the following substantive changes: we used SupT1 cells expressing CCR5 to support replication of viruses with transmitted-founder Envs, we used more virions for the first passage ($\ge 3 \times 10^6$ versus $5\times 10^5$ infectious units per library) to avoid bottlenecking library diversity, and rather than performing a full second passage we just did a short high-MOI infection to enable recovery of \textit{env} genes from infectious virions without bottlenecking (Figure~\ref{fig:dms_schematic}A).
We performed this deep mutational scanning in full biological triplicate for both BG505 and BF520 (Figure~\ref{fig:dms_schematic}B).
Our deep mutational scanning libraries encompassed all codon mutations to all sites in Env except for the signal peptide and cytoplasmic tail.

@@

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

\subsection{Amino-acid preferences of the Envs and their relationship to HIV evolution in nature}
The most immediate question is how authentically the experimental measurements describe the actual selection on Env function in nature.
Direct comparisons between experimentally measured amino-acid preferences and amino-acid frequencies in natural sequences are confounded by the fact that the natural sequences are evolutionarily related.
This problem can be overcome by making the comparison in a phylogenetic context.
Specifically, we used our deep mutational scanning data to construct experimentally informed codon models (ExpCM's) for Env evolution~\citep{hilton2017phydms}.
Importantly, since we expect many sites in Env to be under diversifying selection from immunity, we extended the ExpCM's to draw the relative dN/dS parameter ($\omega$) from a gamma distribution as is commonly done for codon-substitution models~\citep{yang2000codon}.

@@

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BG505_prefs.pdf}}
\caption{\label{fig:BG505prefs}
Amino-acid preferences for the BG505 Env.
At each site, the height of the letter is proportional for that site's preference for that amino acid.
The top color bar indicates the region of Env (gp120 variable loop, gp120 not variable loop, or gp41).
The lower color bar indicates the evidence that the site evolves faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected given the experiments~\citep[see][]{bloom2017identification}.
The letters above the logos indicate the wildtype amino acid in BG505.
Sites are numbered using the HXB2 scheme~\citep{korber1998numbering}).
This logo plot shows the site-specific amino-acid preferences for BG505 after averaging the replicates and re-scaling by the stringency parameter in Table \ref{tab:phydms}.
The figure was generated using \texttt{dms\_tools2}~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}, which in turn utilizes \texttt{weblogo}~\citep{crooks2004weblogo}.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BG505\_prefs.csv.}}
\figdata{The sequence of BG505 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BG505\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BG505 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

\begin{figure}
\begin{fullwidth}
{\includegraphics[width=1.3\textwidth]{figures/BF520_prefs.pdf}}
\caption{\label{fig:BF520prefs}
Amino-acid preferences for the BF520 Env.
This figure is the same as Figure~\ref{fig:BG505prefs} except that it shows the data for BF520 instead of BG505.
}
\figdata{The numerical values of the amino-acid preferences plotted in this figure are in \texttt{rescaled\_BF520\_prefs.csv.}}
\figdata{The sequence of BF520 Env and mapping from sequential (\emph{original} column) to HXB2 numbering (\emph{new} column) is in \texttt{BF520\_to\_HXB2.csv}.}
\figdata{The $\omega_r$ values and associated $P$-values for BF520 in HXB2 numbering are in \texttt{BG505\_omegabysite.tsv}.}
\end{fullwidth}
\end{figure}

@@

The overall distribution of shifts between BG505 and BF520 is shown in Figure~\ref{fig:prefsdist}B.
Most sites have relatively small shifts (close to zero), although there is a long tail of sites with large shifts.
This tail reaches its upper value with site 512, which has a shift of 0.52 out of a maximum possible of 1.0.
How should we interpret this distribution -- have mutational effects shifted a lot, or not very much?
We can establish an upper-bound for how much sites might shift by comparing Env to a \emph{non}-homologous protein.
Figure~\ref{fig:prefsdist}B shows the distribution of shifts when comparing Env to influenza's hemagglutinin protein, which has previously had its amino-acid preferences measured by deep mutational scanning~\citep{doud2016accurate} .
Most sites have large shifts between Env and hemagglutinin, with the typical shift being $\sim$0.4 and some approaching the maximum value of 1.0.
We can also establish a lower-bound by creating a null distribution for the expected shifts if all differences are simply due to experimental noise.
This null distribution is created by randomizing the experimental replicates among Envs.
Figure~\ref{fig:prefsdist}B shows that the null distribution is more peaked at zero than the real distribution, and does not have the same prominent tail of sites with large shifts.
The answer to the question of how much mutational effects have shifted is therefore nuanced: they have substantially shifted at some sites, but remain vastly more similar between the two Envs than between two unrelated proteins.

@@

However, not all of the significantly shifted sites show a simple pattern of broadening or narrowing of mutational tolerance.
For instance, site 288 does not alter its mutational tolerance but rather flips its rather narrow amino-acid preference from phenylalanine in BG505 to leucine in BF520 (Figure~\ref{fig:prefsdist}C).
Several other sites show similar but somewhat milder flips in which amino acids they prefer.
In later sections, we investigate evolutionary and structural characteristics of these shifted sites -- but as is often the case for proteins as complex as Env, we are unable to rationalize many aspects of the observed sequence-function relationships, including the cause of these shifts.

@@

What distinguishes the sites that have undergone significant shifts?
Figure~\ref{fig:shiftfeatures}A shows the locations of the shifted sites on the crystal structure of Env.
There is no visually obvious tendency for shifted sites to preferentially be on Env's surface or in its core, and a statistical analysis (Figure~\ref{fig:shiftfeatures}B) finds no association between a site's relative solvent accessibility and whether its amino-acid preferences have shifted.
However, visual inspection of Figure~\ref{fig:shiftfeatures}A does suggest that the sites of significant shifts tend to cluster in Env's structure, and a statistical analysis confirms that this is the case (Figure~\ref{fig:shiftfeatures}C).
Therefore, the factors that drive shifts in Env's mutational tolerance often affect physically interacting clusters of residues in a coordinated fashion.

An obvious hypothesis is that strongly shifted sites have substituted between BG505 and BF520, or physically contact such substitutions.
According to this hypothesis, substitutions would alter the local physicochemical environment of the substituted site and its neighbors, thereby shifting the amino-acid preferences of sites in the physical cluster.
But surprisingly, the typical magnitude of shifts is not significantly larger at sites that have substituted or contact substitutions compared to other sites in Env (Figure~\ref{fig:shiftfeatures}C).
There is a borderline trend for the significantly shifted sites to be more likely to have substituted between BG505 and BF520 (Figure~\ref{fig:shiftfeatures}-Figure~supplement~\ref{figsupp:subs_shifts}), but most shifted sites have not substituted (only 8 of the 30 shifted sites differ in amino-acid identity between the two Envs).
The lack of strong enrichment in shifts at substituted sites contrasts with previous protein-wide experimental~\citep{doud2015site} and simulation-based~\citep{pollock2012amino,shah2015contingency} studies of shifting amino-acid preferences, which found that shifts were dramatically more pronounced at sites of substitutions.
The difference may arise because these earlier studies examined proteins that are fairly conformationally static (absolutely so in the case of the simulations).
On the other hand, Env is an extremely complex and conformationally dynamic protein~\citep{munro2014conformational,ozorowski2017open}, which may increase the opportunities for long-range epistasis to enable substitutions at one site to shift the amino-acid preferences of distant sites.

@@

\subsection{Comparing selection in the lab to natural selection}
Our experiments measure the effects of mutations on viral replication in a T-cell line in the lab.
But HIV actually evolves in humans, where additional selection pressures on Env are undoubtedly present.
For instance, antibody pressure might increase the rate of evolution at some sites~\citep{albert1990rapid,wei2003antibody,richman2003rapid}, whereas pressure to mask certain epitopes~\citep{kwong2002hiv} might add constraint at other sites. 
Comparing selection in our experiments to natural selection can identify sites that are under such additional pressures during HIV's actual evolution in humans.

@@

We determined whether each site in Env evolves faster or slower in nature than expected given three models: that evolution is purely neutral (all nonsynonymous and synonymous mutations have equivalent effects), that sites are under the protein-level constraint measured in our experiments with BG505, or that sites are under the constraint measured with BF520. 
The first model used a standard dN/dS test~\citep[the ``FEL'' method of][]{kosakovsky2005not}, whereas the other two models performed a conceptually similar test but used ExpCM that incorporate experimental measurements as described by \citet{bloom2017identification}.
The standard dN/dS model finds hundreds of sites evolve slower than expected under neutral evolution, and only a handful of sites evolve faster than expected under neutral evolution (Table~\ref{tab:phydms}).
This finding is unsurprising, since it is well known that Env is under functional constraint and so has many sites that evolve non-neutrally.
In contrast, ExpCM that test the rates of evolution relative to the experimentally measured constraints find far fewer sites that evolve slower than expected, but many more sites that evolve faster (Table~\ref{tab:phydms}).

The sites that evolve slower or faster than expected from the experiments are shown in Figure~\ref{fig:divsel}A,B, and also overlaid on the logoplots in Figures~\ref{fig:BG505prefs} and \ref{fig:BF520prefs} as the $\omega_r$ values.
The identified sites are similar regardless of whether we use the experiments with BG505 or BF520 (Figure~\ref{fig:divsel}C).
The reason the results are similar for both experimental datasets is that (as discussed above) the amino-acid preferences of \emph{most} sites are similar in both Envs, meaning that either dataset provides a reasonable approximation of the site-specific functional constraints across the clade A Envs in Figure~\ref{fig:tree}.

What causes some sites to evolve faster or slower in nature than expected from the experiments?
The answer in both cases is likely to be immune selection.
Most of the sites of faster-than-expected evolution are on the surface of Env (Figure~\ref{fig:divsel}A,B and Figure~\ref{fig:divsel}-Figure~supplement~\ref{figsupp:slower}).
Env's escape from autologous neutralizing antibodies often involves amino-acid substitutions in surface-exposed regions~\citep{moore2009specificity}, including at many of the sites that evolve faster than expected.
Since our deep mutational scanning did not impose antibody pressure, sites where substitutions are antibody-driven will evolve faster in nature than expected from the experiments.
Interestingly, immune selection also offers a plausible explanation for the sites that evolve \emph{slower} than expected.
In addition to escaping immunity via substitutions at antibody-binding footprints, Env is notorious for employing a range of more general strategies to reduce its susceptibility to antibodies.
These strategies include shielding immunogenic regions with glycans~\citep{wei2003antibody,stewart2016trimeric} or hiding them by adopting a closed protein conformation~\citep{kwong2002hiv,guttman2015antibody,ozorowski2017open}.
Sites that contribute to such general immune-evasion strategies will be under a constraint in nature that is not present in our experiments -- and indeed, such sites evolve more slowly than expected from our experiments.
For instance, sites of N-linked glycans change slower than our experiments suggest is required by mere functional selection (Figure~\ref{fig:divsel}C), as do sites that help regulate Env's transition between open and closed conformations that have different antibody susceptibilities (Figure~\ref{fig:divsel}D).
Therefore, we can distinguish evolutionary patterns that are shaped by simple selection for Env function from those that are due to the additional complex pressures imposed during human infections.
